---
title: "Beluga and Narwhal Acoustic Classifier"
author:
  - Marie Zahn
  - School of Aquatic and Fishery Sciences, University of Washington, USA
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

*Click classifiers used in PAMGuard click detector module:*
Unclassified, Code 0 (black dot)
4-20 kHz, Code 1 (orange diamond)
20-50 kHz, Code 2 (red hexagon)
50-70 kHz, Code 3 (blue circle)
70-100 kHz, Code 4 (green star)
100-150 kHz, Code 5 (yellow square)
150-250 kHz, Code 6 (red triangle)

*Data summary:*
Total time: 8:38:55
Total beluga: 1:03:20
Total narwhal: 7:38:15

*PAMpal*
To install the latest version from GitHub, first make sure that you have Rtools installed, then run the following code to install PAMpal.

Visit <https://taikisan21.github.io/PAMpal/> for detailed instructions.

## BANTER classifier

```{r run BANTER}
## load required packages
library(rfPermute)
library(ggplot2)
library(PAMpal)
library(banter)

## install PAMpal package from GitHub
# devtools::install_github('TaikiSan21/PAMpal')

## install BANTER package from GitHub
# devtools::install_github('ericarcher/banter')

## load data - narwhal and beluga echolocation parameters obtained from PAMGuard and PAMpal
load(file='Data/Narwhal-Beluga-data_043021.rdata')

## export banter model
banterAll <- export_banter(Narwhal.Beluga.data)

## alternatively, load banterAll file here:
load(file='Data/banterAll_043021.rdata')

## initialize banter model
bant.mdl <- initBanterModel(banterAll$events)

## run RF models for each Detector added-------------------------------------
bant.mdl <- addBanterDetector(
  bant.mdl, 
  data = banterAll$detectors[c(2,3,4,5)], # does not include Det 1 and 6
  ntree = 10000, 
  sampsize = 50,
  importance = TRUE
)
## look at summary of Detector models
## this shows correct classification rate for each species in each detector
summary(bant.mdl)
plotDetectorTrace(bant.mdl)

## run Event model------------------------------------------------------------
bant.mdl <- runBanterModel(bant.mdl, ntree = 10000, sampsize = 9)
## look at summary for Event model
summary(bant.mdl)

## get RF data from banter model
event.rf <- getBanterModel(bant.mdl, "event")
## get RF data for specific detecto - example of Det 5:
(event.rf.Det5 <- getBanterModel(bant.mdl, "Click_Detector_5"))

## get predictor names for event level RF model
colnames(bant.mdl@model.data)

## can also load banter model results from paper here:
load(file='Data/bantMDL-Narwhal-Beluga-classification_043021.rdata')

## SUMMARIES
summary(bant.mdl)
plotVotes(event.rf)
proximityPlot(event.rf)
casePredictions(event.rf)
```

```{r varImp}
## traditional variable importance plot from rf
varImpPlot(event.rf)
## produces heatmap of variable importance scores
impHeatmap(event.rf)

## Identify most important variables in event model
bant.imp <-  data.frame(importance(event.rf))
bant.4imp <- bant.imp[order(bant.imp$MeanDecreaseAccuracy, decreasing=TRUE),][1:4,] # select top 4 important event stage predictors
bant.4imp # Look at the predictors and use Detector names for next step

## Use most important Detectors identified in previous step to determine which sonar parameters are most important
getDetectorNames(bant.mdl)
bant.det2.rf <- getBanterModel(bant.mdl, "Click_Detector_2")
bant.det3.rf <- getBanterModel(bant.mdl, "Click_Detector_3")

## select and order detector importance data
Det2.imp <- data.frame(importance(bant.det2.rf))
Det3.imp <- data.frame(importance(bant.det3.rf))

Det2.top.imp <- Det2.imp[order(Det2.imp$MeanDecreaseAccuracy, 
                               decreasing=TRUE),3, drop = FALSE]
Det3.top.imp <- Det3.imp[order(Det3.imp$MeanDecreaseAccuracy, 
                               decreasing=TRUE),3, drop = FALSE]
```

***

## Single Random Forests classifier

```{r RF Model}
## load required libraries
library(randomForest)
library(caret)

## You can load RF model object from paper here:
load(file = 'Data/ranfor_classification_043021.rdata')

## Or, you can run the RF model yourself here:

## empty master data frame for entering values in loop below
master_df <- as.data.frame(matrix(nrow = 0, ncol = 25))

events <- Narwhal.Beluga.data@events
nevents <- length(Narwhal.Beluga.data@events)
events_names <- names(events)

## loop to extract desired sonar parameters from PAMpal data (removed PeakHz parameter because it is the same as peak)
for (i in 1:nevents){
  ndetectors <- length(events[[events_names[i]]]@detectors)
  detector_names <- names(events[[events_names[i]]]@detectors)
  detectors <- events[[events_names[i]]]@detectors
  
  for (j in (1:ndetectors)){
    nclicks <- length(detectors[[j]]$eventLabel)

    temp_df <- matrix(nrow = nclicks, ncol = 25) # edit these for number sonar parameters extracted
    colnames(temp_df) <- c("event", "species", "detector", "duration", "peakTime", "peak", "peak2", "peak3", "trough", "trough2", "peakToPeak2", "peakToPeak3", "peak2ToPeak3", "Q_10dB", "Q_3dB", "centerHz_10dB", "centerHz_3dB", "fmin_10dB", "fmax_10dB", "fmin_3dB", "fmax_3dB", "BW_10dB", "BW_3dB", "noiseLevel", "dBPP")

    temp_df[,"event"] <- rep(events_names[i], nclicks)
    temp_df[,"detector"] <- rep(detector_names[j], nclicks)
    temp_df[,"species"] <- detectors[[j]]$eventLabel # narwhal = 085; beluga = 045
    temp_df[,"duration"] <- detectors[[j]]$peak
    temp_df[,"peakTime"] <- detectors[[j]]$peak
    temp_df[,"peak"] <- detectors[[j]]$peak
    temp_df[,"peak2"] <- detectors[[j]]$peak2
    temp_df[,"peak3"] <- detectors[[j]]$peak3
    temp_df[,"trough"] <- detectors[[j]]$trough
    temp_df[,"trough2"] <- detectors[[j]]$trough2
    temp_df[,"peakToPeak2"] <- detectors[[j]]$peakToPeak2
    temp_df[,"peakToPeak3"] <- detectors[[j]]$peakToPeak3
    temp_df[,"peak2ToPeak3"] <- detectors[[j]]$peak2ToPeak3
    temp_df[,"Q_10dB"] <- detectors[[j]]$Q_10dB
    temp_df[,"Q_3dB"] <- detectors[[j]]$Q_3dB
    #temp_df[,"PeakHz_10dB"] <- detectors[[j]]$PeakHz_10dB # supposed to be peak freq as well
    #temp_df[,"PeakHz_3dB"] <- detectors[[j]]$PeakHz_3dB # supposed to be peak freq as well
    temp_df[,"centerHz_10dB"] <- detectors[[j]]$centerHz_10dB
    temp_df[,"centerHz_3dB"] <- detectors[[j]]$centerHz_3dB
    temp_df[,"fmin_10dB"] <- detectors[[j]]$fmin_10dB
    temp_df[,"fmax_10dB"] <- detectors[[j]]$fmax_10dB
    temp_df[,"fmin_3dB"] <- detectors[[j]]$fmin_3dB
    temp_df[,"fmax_3dB"] <- detectors[[j]]$fmax_3dB
    temp_df[,"BW_10dB"] <- detectors[[j]]$BW_10dB
    temp_df[,"BW_3dB"] <- detectors[[j]]$BW_3dB
    temp_df[,"noiseLevel"] <- detectors[[j]]$noiseLevel
    temp_df[,"dBPP"] <- detectors[[j]]$dBPP
    
    master_df <- rbind(master_df, temp_df)
  }
  
  ## change species code to species name for simplicity
  master_df$species[which(master_df$species == "045")] <- "Beluga"
  master_df$species[which(master_df$species == "085")] <- "Narwhal"

}

## make event names a factor
master_df$event <- factor(master_df$event, levels = unique(master_df$event))

params <- c("duration", "peakTime", "peak", "peak2", "peak3", "trough", "trough2", "peakToPeak2", "peakToPeak3", "peak2ToPeak3", "Q_10dB", "Q_3dB", "centerHz_10dB", "centerHz_3dB", "fmin_10dB", "fmax_10dB", "fmin_3dB", "fmax_3dB", "BW_10dB", "BW_3dB", "noiseLevel", "dBPP")

for (i in 1:length(params)){
  master_df[,params[i]] <- as.numeric(master_df[,params[i]])
}

## group data by event and then average all click sonar parameters for each event
Narluga.data <- master_df %>% group_by(event) %>% 
  summarise_all(funs(if(is.numeric(.)) mean(.) else first(.)))

## add ICI to master dataframe
master_ici <- as.data.frame(matrix(nrow = 0, ncol = 1))
for (e in seq_along(events(Narwhal.Beluga.data))){
  # pull out mean ici for the event
  ici <- events[[events_names[e]]]@ancillary[["measures"]][["All_ici"]]
  master_ici <- rbind(master_ici, ici)
}
colnames(master_ici) <- "ici"

## combine ici with master df
Narluga.data <- bind_cols(Narluga.data, master_ici)

## remove event and detector column
Narluga.data.full <- Narluga.data
Narluga.data <- Narluga.data.full[,-c(1,3)]

## Run RANDOM FOREST model -----------------------------------------------------

## make species a factor
Narluga.data$species <- as.factor(Narluga.data$species)
## remove peakTime, dBPP, and noiseLevel
Narluga.data <- Narluga.data %>% select(-c("peakTime", "dBPP", "noiseLevel"))

## Run random forests model
ranfor <- randomForest(species ~ ., 
                       data = Narluga.data, 
                       ntree = 10000, 
                       importance = TRUE,
                       replace = FALSE, 
                       proximity = TRUE, 
                       sampsize = c(9, 9))

ranfor

## Predictions from the ensemble of all the trees
(predict <- ranfor$predicted)

## variables sorted in decreasing order of importance
varImpPlot(ranfor)
proximityPlot(ranfor)

## identify misclassifications ------------------------------------------------
casePredict <- casePredictions(ranfor)

misclass <- casePredict %>% 
  filter(is.correct== FALSE) %>%
  select(case.id)
misclass

Narluga.data.full[misclass$case.id,1:2]

```

```{r RF sensitivity analyses}
## run the RF model over all combinations of mtry and sampsize

## make master df with all possible combinations of mtry and sampsize
mtry <- rep(2:19, each = 17)
sampsize <- rep(2:18, 18)
params <- cbind(mtry, sampsize)

## loop over all possible combinations
master_df <- as.data.frame(matrix(nrow = nrow(params), ncol = 1))
yy <- Narluga.data$species

for (i in 1:nrow(params)){
ranfor <- randomForest(species ~ ., 
                       data = Narluga.data, 
                       ntree = 10000, 
                       importance = TRUE,
                       replace = FALSE, 
                       proximity = TRUE, 
                       sampsize = as.vector(c(params[i,2],params[i,2])),
                       mtry = params[i,1])
## calculate accuracy
rfpred <- predict(ranfor)
accuracy <- caret::confusionMatrix(rfpred, yy)
master_df[i,1] <- accuracy$overall[1]

}

## combine dfs
sensitivity <- cbind(params, master_df)
sensitivity

## calculate % range of model output
max(sensitivity$V1) - min(sensitivity$V1)

```

```{r significance}
rfpred <- predict(ranfor)
yy <- as.factor(Narluga.data$species)
## calculate accuracy and kappa coefficient
caret <- caret::confusionMatrix(rfpred, yy)
caret$overall[1]
```

